---
title: "Predictive Modeling of Neural Activity in Mice"
author: "Ellie Bi"
date: "2024-03-14"
output: html_document
---

# Data Pre-Processing

```{r, warning = FALSE}
library(tidyverse)
library(stats)
library(Rtsne)
library(ROCR)
library(pROC)
library(caret)
library(xgboost)
library(class)
library(e1071)
library(randomForest)
library(nnet)
library(MASS)
```

```{r, warning = FALSE}
setwd("C:/Users/ellie/OneDrive/Documents/rstudio/sta141a/STA141 Project/sessions")

session <- list()
for (i in 1:18) {
  session[[i]] <- readRDS(paste0("session", i, ".rds"))
}
```

```{r}
dataproc <- function(id) {
  trials <- lapply(1:length(session[[id]]$spks), function(trial) {
    spk <- session[[id]]$spks[[trial]]
    spkavg <- matrix(colMeans(spk), nrow = 1)
    colnames(spkavg) <- paste0("bin ", 1:40)
    ba <- session[[id]]$brain_area[which.max(rowMeans(spk))]
    name <- session[[id]]$mouse_name
    
    data.frame(
      SESSID = as.factor(id),
      TRIALID = trial,
      contrast_left = session[[id]]$contrast_left[trial],
      contrast_right = session[[id]]$contrast_right[trial],
      feedback_type = session[[id]]$feedback_type[trial],
      name,
      ba,
      spkavg
    )
  })
  do.call(rbind, trials)
}

totdat <- do.call(rbind, lapply(1:18, dataproc)) # categorical
sesdat <- totdat[, -c(6, 7)] # numerical
```

```{r}
dat <- sesdat[, 6:45]
```


```{r}
head(totdat)
```

# Exploratory Data Analysis

#### Number of Trials

```{r}
num_trials <- sapply(session, function(s) length(s$feedback_type))

for (i in 1:length(num_trials)) {
  cat("Session", i, ":", num_trials[i], "Trials\n")
}
```

### Brain Area

#### Unique Brain Area Names

```{r}
cat("Unique Brain Area Names:\n"); unique(unlist(lapply(session, function(s) unique(s$brain_area))))
```

#### Unique Brain Area Names for Brain Areas that had Most Neural Activity in Each Trial

```{r}
cat("Unique Brain Area Names:\n", unique(totdat$ba))
```

#### Unique Brain Areas per Session

```{r}
un_ba <- sapply(session, function(s) length(unique(s$brain_area)))

for (i in 1:length(un_ba)) {
  cat("Session", i, ":", un_ba[i], "Unique Brain Areas \n")
}
```

```{r}
plt <- data.frame(session = integer(), brain_area = character())

for (i in 1:length(session)) {
  ses <- session[[i]]
  unique_areas <- unique(ses$brain_area)
  
  plt <- rbind(plt, expand.grid(session = i, brain_area = unique_areas))
}

ggplot(plt, aes(x = as.factor(session), y = brain_area)) +
  geom_point() +
  labs(title = "Brain Areas in Each Session",
       x = "Session",
       y = "Brain Area") +
  theme_minimal()
```

#### Brain Area Frequency for Brain Areas that had Most Neural Activity in Each Trial

```{r}
batab <- table(totdat$ba)
badf <- as.data.frame(batab)
badf <- badf[order(badf$Freq, decreasing = TRUE), ] 

ggplot(badf, aes(x = reorder(Var1, Freq), y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Brain Area Frequencies", x = "Brain Area", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_flip()  
```

### Neural Activity

#### Neural Activity Across Sessions

```{r cache = TRUE}
plot(1, type = "n", 
     xlab = 'Time', 
     ylab = 'Neural Activity (spks)', 
     main = 'Nueral Activity Across Sessions', 
     xlim = c(1, 40),  
     ylim = c(1, 20000))

for (i in seq_along(session)) {
  ses <- session[[i]]
  tot_spks <- Reduce('+', ses$spks)
  lines(1:40, colSums(tot_spks), col = rainbow(length(session))[i])
}

legend("topright", 
       legend = paste("Session", 1:length(session)), 
       col = rainbow(length(session)), 
       lty = 1, 
       cex = 0.4)
```

```{r}
plot(colMeans(dat), type = "l", xlab = "Time Bin", ylab = "Average Spike Rate",
     main = "Average Spike Rate Over Time Bins")
```


#### Average Spike Rate Across Trials

```{r warning = FALSE}
avgspk <- sesdat %>%
  summarise(avgspks = rowSums(sesdat[, 6:45]) / 40)

avgspk$SESSID <- sesdat$SESSID
avgspk$TRIALID <- sesdat$TRIALID
avgspk$ba <- totdat$ba
avgspk$name <- totdat$name
```

```{r}
ggplot(avgspk, aes(x = TRIALID, y = avgspks)) + 
  geom_line() +
  geom_smooth(method = "loess") +
  facet_wrap(~SESSID, ncol = 6, nrow = 3) + 
  labs(
    title = "Average Spike Rate Across Trials for Each Session",  
    x = "Session",                                                
    y = "Average Spike Rate"                                   
  ) +                                              
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

#### Average Spike Rate for Each Brain Area

```{r warning = FALSE}
ggplot(avgspk, aes(x = TRIALID, y = avgspks)) + 
  geom_line() +
  geom_smooth(method = "loess") +
  facet_wrap(~ ba) + 
  labs(
    title = "Average Spike Rate Across Brain Area",  
    x = "Brain Area",                                                
    y = "Average Spike Rate"                                   
  ) 
```

#### Average Spike Rate for Each Mice

```{r}
ggplot(avgspk, aes(x = TRIALID, y = avgspks)) + 
  geom_line() +
  geom_smooth(method = "loess") +
  facet_wrap(~ name) + 
  labs(
    title = "Average Spike Rate Across Mice",  
    x = "Mice",                                                
    y = "Average Spike Rate"                                   
  ) 
```

### Contrast Difference

#### Percentage of Contrast Differences for Each Mice

```{r}
totdat$contrast_diff <- abs(totdat$contrast_left - totdat$contrast_right)

counts.df <- totdat[c('name', 'contrast_diff')]
counts.df$contrast_diff <- as.factor(counts.df$contrast_diff)
counts <- table(counts.df)

(percentages <- prop.table(counts, margin = 1))
```

#### Success Rate Correlated with Contrast Differences

```{r}
totdat %>% group_by(contrast_diff) %>% summarize(success = mean(feedback_type == 1, na.rm = TRUE))
```

### Success Rate

```{r}
feedback <- table(unlist(lapply(session, function(s) s$feedback_type)))

for (i in 1:length(feedback)) {
  label <- ifelse(i == 1, "Failure", "Success")
  cat("Feedback Type", i, ":", label, "-", feedback[i], "Occurrences\n")
}
```

```{r}
totdat$tgr = cut(totdat$TRIALID, breaks = seq(0, max(totdat$TRIALID), by = 25),include.lowest = TRUE)
levels(totdat$tgr) <- seq(0, max(totdat$TRIALID), by = 25)[2:18]
totdat$success <- as.numeric(totdat$feedback_type == 1)
```

#### Success Rate of Change over Each Session

```{r}
success <- aggregate(success ~ SESSID + tgr, data = totdat, FUN = function(x) mean(x))

ggplot(success, aes(x = tgr, y = success)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~SESSID, ncol = 3) +
  theme_bw()
```

#### Success Rate of Change over Brain Area

```{r}
success <- aggregate(success ~ ba + tgr, data = totdat, FUN = function(x) mean(x))

ggplot(success, aes(x = tgr, y = success)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ ba) +
      theme_bw()
```

#### Success Rate of Change over Each Mice

```{r}
success <- aggregate(success ~ name + tgr, data = totdat, FUN = function(x) mean(x))

ggplot(success, aes(x = tgr, y = success)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ name) +
      theme_bw()
```

# Clustering

```{r}
dat <- sesdat[, 6:45]
```


### PCA 

```{r}
pca <- prcomp(dat)

pca.df <- as.data.frame(pca$x)
pca.df$SESSID <- totdat$SESSID
pca.df$ba <- totdat$ba
pca.df$name <- totdat$name
```

#### PCA Plot for Session

```{r}
ggplot(pca.df, aes(x = PC1, y = PC2, color = SESSID)) +
  geom_point() +
  labs(title = "PCA: PC1 vs PC2")
```

#### PCA Plot for Brain Area

```{r}
ggplot(pca.df, aes(x = PC1, y = PC2, color = ba)) +
  geom_point() +
  labs(title = "PCA: PC1 vs PC2")
```

#### PCA Plot for Mice

```{r}
ggplot(pca.df, aes(x = PC1, y = PC2, color = name)) +
  geom_point() +
  labs(title = "PCA: PC1 vs PC2")
```

### t-SNE

```{r}
tsne <- Rtsne::Rtsne(dat)
plot(tsne$Y, col = "blue")
```

# Data Integration

### Unscaled Total Train & Test

```{r}
set.seed(825)
idx <- sample(seq_len(nrow(sesdat)), 0.8 * nrow(sesdat))
train <- sesdat[idx, ]
test <- sesdat[-idx, ]
```

```{r}
traindat <- train[, -c(1, 2)]
testdat <- test[, -c(1, 2)]
```

### Unscaled First 20 Bins Train & Test

```{r}
f.traindat <- traindat[, -c(24:43)]
f.testdat <- testdat[, -c(24:43)]
```

### Unscaled Last 20 Bins Train & Test

```{r}
l.traindat <- traindat[, -c(4:23)]
l.testdat <- testdat[, -c(4:23)]
```

### Scaled Total Train & Test

```{r}
s.traindat <- train[, -c(1, 2)]
s.testdat <- test[, -c(1, 2)]

s.traindat[, c(4:43)] <- scale(s.traindat[, c(4:43)])
s.testdat[, c(4:43)] <- scale(s.testdat[, c(4:43)])
```

### Scaled First 20 Bins Train & Test

```{r}
s.f.traindat <- s.traindat[, -c(24:43)]
s.f.testdat <- s.testdat[, -c(24:43)]
```

### Scaled Last 20 Bins Train & Test

```{r}
s.l.traindat <- s.traindat[, -c(4:23)]
s.l.testdat <- s.testdat[, -c(4:23)]
```

### Formulas:

#### F1 Score

```{r}
f1 <- function(confmat) {
  TP <- confmat[2, 2]
  FP <- confmat[1, 2]
  FN <- confmat[2, 1]
  
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  
  f1_score <- 2 * precision * recall / (precision + recall)
  
  return(f1_score)
}
```

# Model Building

### XGBoost

```{r}
xgb <- function(train, test) {
  # model building
  xgbtrain <- ifelse(train$feedback_type == 1, 1, 0)
  
  xgbmat <- xgb.DMatrix(data = as.matrix(train[, -which(names(train) %in% c("feedback_type"))]), 
                       label = xgbtrain)

  xgbmodel <- xgboost(data = xgbmat, 
                      objective = 'binary:logistic', 
                      eval_metric = 'error',
                      nrounds = 10)
  
  # predictions
  xgbpred <- predict(xgbmodel, as.matrix(test[, -which(names(test) %in% c("feedback_type"))]))
  binary <- ifelse(xgbpred > 0.5, 1, -1)
  
  # data
  cat('\nFive Number Summary of Predictions\n')
  print(summary(binary))
  
  # confusion matrix
  xgb.confmat <- table(Actual = test$feedback_type, Predicted = binary)
  cat('\n Confusion Matrix\n')
  print(xgb.confmat)
  
  # accuracy
  xgb.acc <- sum(diag(xgb.confmat)) / sum(xgb.confmat)
  cat(paste("\nAccuracy for XGBoost:", round(xgb.acc, 4), '\n'))
  
  # f1
  xgb.f1 <- f1(xgb.confmat)
  cat(paste("\nF1 Score for XGBoost:", round(xgb.f1, 4), '\n'))
  
  
  # AUROC
  auroc <- auc(roc(test$feedback_type, xgbpred))
  cat(paste("\nArea under ROC for XGBoost:", round(auroc, 4)))
}
```

#### XGBoost: Total Data

```{r}
xgb(traindat, testdat)
```

#### XGBoost: First Bins

```{r}
xgb(f.traindat, f.testdat)
```

#### XGBoost: Last Bins

```{r}
xgb(l.traindat, l.testdat)
```

### Random Forest

```{r}
rf <- function(train, test) {
  # model building
  predictors <- setdiff(names(train), "feedback_type")

  rfmodel <- randomForest(
    formula = as.formula("feedback_type ~ ."),
    data = train,
    ntree = 500,
    mtry = sqrt(length(predictors)),
    importance = TRUE
  )

  # predictions
  rfpred <- predict(rfmodel, newdata = test[, -which(names(test) %in% c("feedback_type"))])
  binary <- ifelse(rfpred > 0, 1, -1)

  # data
  cat('\nFive Number Summary of Predictions\n')
  print(summary(binary))

  # confusion matrix
  rf.confmat <- table(Actual = test$feedback_type, Predicted = binary)
  cat('\n Confusion Matrix\n')
  print(rf.confmat)
  
  # accuracy
  rf.acc <- sum(diag(rf.confmat)) / sum(rf.confmat)
  cat(paste("\nAccuracy for Random Forest:", round(rf.acc, 4), '\n'))

  # f1
  rf.f1 <- f1(rf.confmat)
  cat(paste("\nF1 Score for Random Forest:", round(rf.f1, 4), '\n'))

  # AUROC
  auroc <- auc(roc(test$feedback_type, rfpred))
  cat(paste("\nArea under ROC for Random Forest:", round(auroc, 4)))
}
```

#### Random Forest: Total Data

```{r warning = FALSE}
rf(traindat, testdat)
```

#### Random Forest: First Bins

```{r warning = FALSE}
rf(f.traindat, f.testdat)
```

#### Random Forest: Last Bins

```{r warning = FALSE}
rf(l.traindat, l.testdat)
```

### Neural Network

```{r}
nn <- function(train, test) {
  # model building
  nnmodel <- nnet(feedback_type ~ ., data = train, size = 5, linout = TRUE) 
  
  # Predictions
  nnpred <- predict(nnmodel, newdata = test[, -which(names(test) %in% c("feedback_type"))], type = "raw")
  binary <- ifelse(nnpred > 0, 1, -1) 
  
  # data
  cat('\nFive Number Summary of Predictions\n')
  print(summary(binary))
  
  # confusion matrix
  nn.confmat <- table(Actual = test$feedback_type, Predicted = binary)
  cat('\nConfusion Matrix\n')
  print(nn.confmat)
  
  # accuracy
  nn.acc <- sum(diag(nn.confmat)) / sum(nn.confmat)
  cat(paste("\nAccuracy for Neural Network:", round(nn.acc, 4), '\n'))
  
  # F1 Score
  nn.f1 <- f1(nn.confmat)
  cat(paste("\nF1 Score for Neural Network:", round(nn.f1, 4), '\n'))
  
  # AUROC
  auroc <- auc(roc(test$feedback_type, nnpred))
  cat(paste("\nArea under ROC for Neural Network:", round(auroc, 4)))
}
```

#### Neural Network: Total Data

```{r}
nn(s.traindat, s.testdat)
```

#### Neural Network: First Bins

```{r warning = FALSE}
nn(s.f.traindat, s.f.testdat)
```

#### Neural Network: Last Bins

```{r warning = FALSE}
nn(s.l.traindat, s.l.testdat)
```

## Final Chosen Model - Random Forest for Total Data

# Test Data Preparation

```{r}
dataproc_test <- function(sessiondat) {
  trials <- lapply(1:length(sessiondat$spks), function(trial) {
    spk <- sessiondat$spks[[trial]]
    spkavg <- matrix(colMeans(spk), nrow = 1)
    colnames(spkavg) <- paste0("bin ", 1:40)
    
    data.frame(
      contrast_left = sessiondat$contrast_left[trial],
      contrast_right = sessiondat$contrast_right[trial],
      feedback_type = sessiondat$feedback_type[trial],
      spkavg
    )
  })
  do.call(rbind, trials)
}
```

```{r}
rfp <- function(test) {
  # model building using rf total data model
  predictors <- setdiff(names(traindat), "feedback_type") 
  
  rfmodel <- randomForest(
    formula = as.formula("feedback_type ~ ."),
    data = traindat,
    ntree = 500,
    mtry = sqrt(length(predictors)),
    importance = TRUE
  )

  # predictions
  rfpred <- predict(rfmodel, newdata = test[, -which(names(test) %in% c("feedback_type"))])
  binary <- ifelse(rfpred > 0, 1, -1)

  # data
  cat('\nFive Number Summary of Predictions\n')
  print(summary(binary))

  # confusion matrix
  rf.confmat <- table(Actual = test$feedback_type, Predicted = factor(binary, levels = c(-1, 1)), exclude = NULL)
  cat('\n Confusion Matrix\n')
  print(rf.confmat)
  
  # accuracy
  rf.acc <- sum(diag(rf.confmat)) / sum(rf.confmat)
  cat(paste("\nAccuracy for Random Forest:", round(rf.acc, 4), '\n'))

  # f1
  rf.f1 <- f1(rf.confmat)
  cat(paste("\nF1 Score for Random Forest:", round(rf.f1, 4), '\n'))

  # AUROC
  auroc <- auc(roc(test$feedback_type, rfpred))
  cat(paste("\nArea under ROC for Random Forest:", round(auroc, 4)))
}
```

# Test Data

```{r warning = FALSE}
setwd("C:/Users/ellie/OneDrive/Documents/rstudio/sta141a/STA141 Project")

test1 <- readRDS("test1.rds")
test2 <- readRDS("test2.rds")
```

```{r}
test1dat <- dataproc_test(test1)
test2dat <- dataproc_test(test2)
```

```{r}
rfp(test1dat)
```

```{r}
rfp(test2dat)
```







